---
title: "Bayesian Linear Analysis"
author: "Megan Feddern"
date: "2023-04-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(nord)
library(tidyr)
library(lubridate)
library(readr)
library(stringr)
library(ggplot2)
library(cowplot)
library(Cairo)
library(MCMCvis)
library(HDInterval)
library(reshape2)
library(tidyverse)
library(dplyr)
library(rstan)
library(here)

```

## Combining Bakun Data
1) Import the Bakun upwelling data from each individual file associated with the 13 lat/long locations. Data versions are: 1˚ 6-hourly version

```{r combine, include=FALSE}
here::i_am("AtmosClimateProj/Src/BayesianLinearModels.Rmd")
bakundat <-read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI246hr_d68d_e898_8529.csv'))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI276hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI306hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI336hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI366hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI396hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI426hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI456hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI486hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI516hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI546hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI576hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI606hr_d68d_e898_8529.csv')))%>%
  bind_rows(read.csv(here('AtmosClimateProj/data/physical/Bakun/erdUI616hr_d68d_e898_8529.csv')))
```

2) Extract yyyy-mm-ddThh:mm:ss from the 'time' column and add region indicator based on station ID.

```{r dates, include=FALSE}
bakun <- bakundat%>%
add_column('Year'=as.numeric(format(as.Date(bakundat$time),"%Y")))%>%
add_column('Month'=as.numeric(format(as.Date(bakundat$time),"%m")))%>%
add_column("Day"=as.numeric(format(as.Date(bakundat$time),"%d")))

bakun_region <- bakun%>%
  filter(station_id=='33N'|station_id=='36N'|station_id=='39N')%>%
  mutate(region="Southern CC")%>%
bind_rows(bakun%>%
  filter(station_id=='42N'|station_id=='45N'|station_id=='48N'|station_id=='51N')%>%
  mutate(region="Northern CC"))%>%
bind_rows(bakun%>%
  filter(station_id=='54N'|station_id=='57N'|station_id=='60N')%>%
  mutate(region="GoA"))


```
3) Summarize across days and hours to get monthly means for region and year. If you revert to seasonal summaries return to here to summarise across dates without taking means of means.
```{r summ, include=FALSE}
bakun_summ <- bakun_region%>%
  group_by(region, Year, Month) %>%
  summarise(monthly_mean = mean(upwelling_index), monthly_sd = sd(upwelling_index))
```
4) Import PDO data
```{r PDO, include=FALSE}
bakun_summ <- bakun_region%>%
  group_by(region, Year, Month) %>%
  summarise(monthly_mean = mean(upwelling_index), monthly_sd = sd(upwelling_index))
```



## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
