---
title: "SOM examples -- spring SST"
format: html
editor: visual
execute: 
  echo: true
  warning: false
  message: false
---

### SOM Setup

```{r}
#| echo: false
#| warning: false
#| message: false
library(ncdf4)
library(chron)
library(tidyverse)
library(kohonen) # fitting
library(aweSOM) # plotting
library(SOMbrero) # plotting
library(paletteer) #colors
library(PNWColors) #more colors
library(here) #navigating folders
set.seed(1234)
here::i_am("som/sst_soms_spring.qmd")
```

```{r loaddata}
#| echo: false
#| warning: false
#| message: false
# start by loading NE Pacific SST
nc <- nc_open("som/copernicus_jun27.nc")
```

```{r getcoords}
#| echo: false
#| warning: false
#| message: false
# get lat/long
x <- ncvar_get(nc, "longitude")
y <- ncvar_get(nc, "latitude")
lat <- rep(y, length(x))   # Vector of latitudes
lon <- rep(x, each = length(y))   # Vector of longitudes
```

```{r gettimes}
#| echo: false
#| warning: false
#| message: false
# get times
raw <- ncvar_get(nc, "time")
tunits<-ncatt_get(nc,"time",attname="units")
tustr<-strsplit(tunits$value, " ")
dates <- RNetCDF::utcal.nc("hours since 1900-01-01 00:00:00.0", raw)
dates <- as.data.frame(dates[,c("year","month")])
dates$index <- seq(1, nrow(dates))
dates$spring_year <- dates$year
dates$spring_year[which(dates$month %in% 7:12)] <- NA
dates$spring_year[which(dates$month %in% 1:3)] <- NA # incremenent 
```

Now we can extract the variable of interest

```{r}
var_name = "sst"
```
Determining the dimensions (nodes)

```{r}
dim <- c(3,2)
```

```{r extract}
#| echo: false
#| warning: false
#| message: false
tmp_array <- ncvar_get(nc,var_name)
dlname <- ncatt_get(nc,var_name,"long_name")
dunits <- ncatt_get(nc,var_name,"units")
fillvalue <- ncatt_get(nc,var_name,"_FillValue")
# replace netCDF fill values with NA's
tmp_array[tmp_array==fillvalue$value] <- NA
X <- tmp_array
```

```{r process}
#| echo: false
#| warning: false
#| message: false
# The 1st dimension of X is lon, 2nd dimension is lat, 3rd dimension is date. But the array needs to be flattened (time on rows, spatial cells on columns) for easier use.
X <- aperm(X, 3:1) # transpose array # X
X <- matrix(X, nrow=dim(X)[1], ncol=prod(dim(X)[2:3])) # months in rows, cells in columns! 
to_drop <- which(is.na(apply(X,2,sum)))
X <- X[,-to_drop]# drop cells with NAs 
spring_lat <- lat[-to_drop]
spring_lon <- lon[-to_drop]
# Add block for spring average calculations
spring_years <- unique(dates$spring_year)
spring_years <- sort(spring_years[-which(is.na(spring_years))])
X_spring <- matrix(NA, length(spring_years), ncol(X))
for(i in 1:length(spring_years)) {
  X_spring[i,] = colMeans(X[which(dates$spring_year == spring_years[i]),])
}

```

### Fitting the SOM model

```{r}
#| echo: false
#| warning: false
#| message: false
grid_dim1 <- 3
grid_dim2 <- 2 
grid_shape <- 'rectangular' 

# fit <- som(scale(X_spring), 
#                grid=somgrid(grid_dim1, grid_dim2, grid_shape))

#fit <- trainSOM(x.data = scale(X_spring), dimension = c(grid_dim1,grid_dim2), verbose = FALSE, nb.save = 0, topo = "square")
```

But because of the sensitivity to grids, we should iterate over grids of varying size. Here we'll just start with a 4x4, which should be large enough?

```{r}
#| echo: false
#| warning: false
#| message: false
grids <- expand.grid(xdim = 2:7)
grids$ydim <- grids$xdim + 1
grids <- rbind(grids, data.frame(xdim = 2:7, ydim = 2:7))
grids$mean_dist <- NA
grids$size <- grids$xdim * grids$ydim
# fit <- list()
# for(i in 1:nrow(grids)) {
#   # kohonen
#   #fit[[i]] <- som(scale(X_spring), 
#   #             grid=somgrid(grids$xdim[i], grids$ydim[i], grid_shape))
#   #grids$mean_dist[i] <- mean(fit[[i]]$distances)
#   # SOMbrero
#   fit[[i]] <- trainSOM(x.data = scale(X_spring), dimension = c(grids$xdim[i],grids$ydim[i]), verbose = FALSE, nb.save = 0, topo = "square")
# }

fit <- trainSOM(x.data = scale(X_spring), dimension = dim, verbose = FALSE, nb.save = 0, topo = "square")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#plot(grids$size, grids$mean_dist, xlab="Grid size", ylab="Mean distance")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#best_model <- which(grids$size == 16)

output <- data.frame(spring_year = spring_years,
                     cluster = fit$clustering)

knitr::kable(output)

```

```{r}
#| echo: false
#| warning: false
#| message: false
X_spring <- as.data.frame(X_spring)
X_spring$year <- spring_years

# convert wide to long
spring_long <- pivot_longer(X_spring, cols = 1:(ncol(X_spring)-1))
spring_long$cell_id <- as.numeric(substr(spring_long$name,2,length(spring_long$name)))
spring_long$lat <- spring_lat[spring_long$cell_id]
spring_long$long <- spring_lon[spring_long$cell_id]
spring_long <- dplyr::rename(spring_long, spring_year = year) 
# join in cluster IDs
spring_long <- dplyr::left_join(spring_long, output)

# Now calculated cluster avgs
cluster_avg <- dplyr::group_by(spring_long, cell_id) %>%
  dplyr::mutate(z = (value - mean(value))/sd(value)) %>%
  dplyr::group_by(cell_id, cluster) %>%
  dplyr::summarise(mean = mean(z),
                   lat = lat[1],
                   long = long[1])

```

```{r}
#| echo: false
#| warning: false
#| message: false
g = ggplot(cluster_avg, aes(long, lat)) + 
  geom_raster(aes(fill = mean)) + 
  facet_wrap(~cluster) + 
  #scale_color_gradient2() + 
  scale_fill_gradient2(low = "blue", high = "red") + 
  xlab("Lon") + 
  ylab("Lat")

g
```

```{r}
#| echo: false
#| warning: false
#| message: false
# png("plots/spring_map_sst.png")
# g
# dev.off()
```

Interpreting this many of SOM prototypes is challenging, and many of the maps appear to be redundant. One approach used with SOMs is post-hoc clustering of the maps. With spring SST, there appears to be \~ 4 groups:

```{r}
#| echo: false
#| warning: false
#| message: false
plot(superClass(fit))

```


```{r}
#| echo: false
#| warning: false
#| message: false
sc <- superClass(fit, k = 4)
output_cluster <- data.frame(cluster = seq(1,length(sc$cluster)), super_cluster = sc$cluster)
knitr::kable(output_cluster)


output <- output%>%dplyr::left_join(output_cluster)

ggplot(output, aes(spring_year, cluster)) + 
  geom_point()
```

```{r}
#| echo: false
#| warning: false
#| message: false
plot(sc, what = "obs", type = "hitmap", maxsize = 20)
```

```{r}
#| echo: false
#| warning: false
#| message: false
col<-pnw_palette("Sunset2",3,type="discrete")
climate_dat <-readRDS(here('data/physical/climate_dat.rds'))%>%
  select(Year_lag, season, period, period2,
          seasonal_PDO, seasonal_NPGO, seasonal_ONI)%>%
  distinct()%>%
  filter(Year_lag<2023)%>%
  rename(spring_year=Year_lag, NPGO = seasonal_NPGO, PDO = seasonal_PDO, ENSO= seasonal_ONI)%>%
  dplyr::left_join(output, by = 'spring_year')%>%
 pivot_longer(
  cols = c(PDO, ENSO, NPGO),
  names_to = c("Index"),
  values_to = "value")%>% 
  filter(season=='Spring') 


ggplot(climate_dat, aes(x = cluster, y=value,  label=spring_year, aes(col=period))) +
  facet_wrap(.~Index, ncol = 3, scales='free') +
  geom_text(aes(col=period))+
  scale_color_manual(values =  col[1:3], labels=c('1940 - 1988', '1989 - 2013', '2014 - 2022'))+
  ylab("Index Value")
```

```{r}
#| echo: false
#| warning: false
#| message: false
ggplot(climate_dat, aes(x = cluster, y=value,  label=spring_year, aes(col=period2))) +
  facet_wrap(.~Index, ncol = 3, scales='free') +
  geom_text(aes(col=period2))+
  scale_color_manual(values =  col[1:2], labels=c('1940 - 1979', '1979 - 2022'))+
  ylab("Index Value")

```

\break

### Alternative standardization

In the above example, we scaled the time series in each spatial cell to be Normal(0,1). That removes the spatial pattern, but not the global trend or the spatial trend. We can re-run the above models using each of these approaches.

#### Standardize + remove global trend

```{r}
#| echo: false
#| warning: false
#| message: false

X_spring <- matrix(NA, length(spring_years), ncol(X))
for(i in 1:length(spring_years)) {
  X_spring[i,] = colMeans(X[which(dates$spring_year == spring_years[i]),])
}

# remove annual mean for each year
for(i in 1:nrow(X_spring)) {
  X_spring[i,] = X_spring[i,] - mean(X_spring[i,], na.rm=T)
}
X_spring <- scale(X_spring)

grids <- expand.grid(xdim = 2:7)
grids$ydim <- grids$xdim + 1
grids <- rbind(grids, data.frame(xdim = 2:7, ydim = 2:7))
grids$mean_dist <- NA
grids$size <- grids$xdim * grids$ydim
# fit <- list()
# for(i in 1:nrow(grids)) {
#   fit[[i]] <- som(scale(X_spring), 
#                grid=somgrid(grids$xdim[i], grids$ydim[i], grid_shape))
#   grids$mean_dist[i] <- mean(fit[[i]]$distances)
# }

fit <- trainSOM(x.data = X_spring, dimension = dim, verbose = FALSE, nb.save = 0, topo = "square")
```

```{r}
#| echo: false
#| warning: false
#| message: false
#best_model <- which(grids$size == 16)

output <- data.frame(spring_year = spring_years,
                     cluster = fit$clustering)

knitr::kable(output)

ggplot(output, aes(spring_year, cluster)) + 
  geom_point()
```

```{r}
#| echo: false
#| warning: false
#| message: false
X_spring <- as.data.frame(X_spring)
X_spring$year <- spring_years

# convert wide to long
spring_long <- pivot_longer(X_spring, cols = 1:(ncol(X_spring)-1))
spring_long$cell_id <- as.numeric(substr(spring_long$name,2,length(spring_long$name)))
spring_long$lat <- spring_lat[spring_long$cell_id]
spring_long$long <- spring_lon[spring_long$cell_id]
spring_long <- dplyr::rename(spring_long, spring_year = year) 
# join in cluster IDs
spring_long <- dplyr::left_join(spring_long, output)

# Now calculated cluster avgs
cluster_avg <- dplyr::group_by(spring_long, cell_id) %>%
  dplyr::mutate(z = (value - mean(value))/sd(value)) %>%
  dplyr::group_by(cell_id, cluster) %>%
  dplyr::summarise(mean = mean(z),
                   lat = lat[1],
                   long = long[1])

```

```{r}
#| echo: false
#| warning: false
#| message: false
g = ggplot(cluster_avg, aes(long, lat)) + 
  geom_raster(aes(fill = mean)) + 
  facet_wrap(~cluster) + 
  #scale_color_gradient2() + 
  scale_fill_gradient2(low = "blue", high = "red") + 
  xlab("Lon") + 
  ylab("Lat")

g
```

```{r}
#| echo: false
#| warning: false
#| message: false
plot(superClass(fit))
```

```{r}
#| echo: false
#| warning: false
#| message: false
sc <- superClass(fit, k = 4)
output_cluster <- data.frame(cluster = seq(1,length(sc$cluster)), super_cluster = sc$cluster)
knitr::kable(output_cluster)

output <- output%>%dplyr::left_join(output_cluster)
```

```{r}
#| echo: false
#| warning: false
#| message: false
#ggplot(output, aes(spring_year, super_cluster)) + geom_point()

ggplot(output, aes(x = spring_year, y=1, fill=as.factor(super_cluster))) +
  geom_tile()+
  scale_fill_manual(values=paletteer_c("grDevices::Set 3", 4))+
   theme_classic()+
  xlab("Spring SST")+
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
         axis.title.y=element_blank(),
        axis.line.y=element_blank())
```

```{r}
#| echo: false
#| warning: false
#| message: false
plot(sc, what = "obs", type = "hitmap", maxsize = 20)
```

```{r}
#| echo: false
#| warning: false
#| message: false
col<-pnw_palette("Sunset2",3,type="discrete")
climate_dat <-readRDS(here('data/physical/climate_dat.rds'))%>%
  select(Year_lag, season, period, period2,
          seasonal_PDO, seasonal_NPGO, seasonal_ONI)%>%
  distinct()%>%
  filter(Year_lag<2023)%>%
  rename(spring_year=Year_lag, NPGO = seasonal_NPGO, PDO = seasonal_PDO, ENSO= seasonal_ONI)%>%
  dplyr::left_join(output, by = 'spring_year')%>%
 pivot_longer(
  cols = c(PDO, ENSO, NPGO),
  names_to = c("Index"),
  values_to = "value")%>% 
  filter(season=='Spring') 


ggplot(climate_dat, aes(x = cluster, y=value,  label=spring_year, aes(col=period))) +
  facet_wrap(.~Index, ncol = 3, scales='free') +
  geom_text(aes(col=period))+
  scale_color_manual(values =  col[1:3], labels=c('1940 - 1988', '1989 - 2013', '2014 - 2022'))+
  ylab("Index Value")

ggplot(climate_dat, aes(x = cluster, y=value,  label=spring_year, aes(col=period2))) +
  facet_wrap(.~Index, ncol = 3, scales='free') +
  geom_text(aes(col=period2))+
  scale_color_manual(values =  col[1:2], labels=c('1940 - 1979', '1979 - - 2022'))+
  ylab("Index Value")

```
